Verifying Hierarchical Cache Coherence Protocols: a theorem-prover based modular, reusable proof

### 2014-02-04 paper/Background.tex
System is a little more motivated. Pull out bits of Proof structure to end and drop into new section
stuff

\section{Correctness of our Cache Coherence System}

Before we can discuss our protocol framework and it's verification we
must first establish exactly what it means for a system to be
correct.

Intuitively, the easiest view of shared memory is as a single ported
memory without caches multiplexed amongst the many individual clients/processors. In
this view, all operations in the system appear to have a single unique
total ordering.

All modern systems have caches between the processor and memory for faster memory access.
It is often also useful for performance to permit some reorderings fo requests from a processor to the memory.
For instance, in the context of a processor with local store buffer, we may
wish to allow store values to bypass subsequent load requests. The specification of
 which reorderings are permitted and which are restricted defines the memory consistency
model.
%, though they may result in a stronger verification requirement.

%As part of our design exploration for our concrete protocol example,
%we may want to change the memory consistency model. To this end, we
%are using the notion of store-atomicity to define our memory
%model. Most memory consistency models require that the memory
%subsystem obeys the store-atomicity property and as such our design
%should be modifiable to any of these designs.

Most memory consistency models can be partitioned into two
parts~\cite{Arvind-memory-model}: a), a ``global'' store-atomicity
property stating that for each location/address in memory, every read and write access to that 
location occurs in some total order with each operation
seeing the effects of previous operations and not later ones (each location behaves like an atomic register) and b),
a ``local'' reordering restriction between requests to different
locations within each processor. The advantage of this approach is
that reordering restrictions pertain to reordering requests locally within
the same processor, while the store-atomicity property is across the whole
system. Thus changing the memory model
by modifying the reordering restrictions will not incur a reverification of
the store-atomicity property.

%For our concrete
%protocol example, we will assume that our memory model is weakly
%ordered, in that there are inter-address dependencies between
%processors.

%% For example, it dictates whether a store request to an address $x$ can
%% be issued to the memory sub-system before a load request to a
%% different address $y$, which was generated by the same processor
%% earlier (\ie load $y$ happened before store $x$ is program order, but
%% the order of issue to memory sub-system is reversed).

The mechanism employed by the memory sub-system to ensure store-atomicity per
location, in the presence of caches, is known as \emph{cache coherence}. In the
absence of caches, since there is exactly one copy of the data for each
address, it is easy to ensure cache coherence. The problem comes in the
presence of shared caches -- multiple copies of the same address can exist in
different caches, and if these locations are independently written, then the
same address can end up having different values in two different caches, and
each location may not behave like an atomic register unless some sort of
co-ordination happens between the different caches -- this co-ordination is
summed up as the \emph{cache coherence protocol}.

Now that we have informally discussed the meaning of cache coherence and store
atomicity, we can now provide a formal definition.

\subsection{Formal definition of store atomicity}

To properly define the interface with between the processor and the
caching system we must define the request and responses for each
processor. 
The requests issued by the processors to the memory sub-system can be modeled
using a function \reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
processor $p$.
The processor name and the position of the request within the processor
uniquely identifies the request.

Figure~\ref{req-resp} shows the
data-type corresponding to a request issued from the processor to the
memory sub-system.  A request contains the address (\addrQ)
corresponding to the request, and a type indicating whether the
request is a load or a store (\desc). In case of a store request, it
also contains data (\dataQ) supplied by the processor for that
address. A response contains a label
(\labelR) to identify the request corresponding to the response. The label
contains a processor name and a position, which uniquely identifies a request.
In the case of a load request, the memory sub-system returns \dataR, the data
returned by the memory sub-system for that address. The memory sub-system also
returns a natural number, \timeR, which is used solely in the definition of
store atomicity, and is not returned in a real system. \timeR{} can be informally
thought to represent the position at which this request has been handled
globally among all the requests issued from all the processors.

@@

\begin{figure}
\centering
\begin{subfigure}{5.5cm}
\begin{tabular}{|lp{4.5cm}|}
\hline
\multicolumn{2}{|c|}{\Request}\\
\hline
\addrQ:& Address associated with the request\\
\desc:& Request type (\Ld{} or \St)\\
\dataQ:& Data associated with a \St{} request\\
\hline
\end{tabular}
\end{subfigure}~~~~
\begin{subfigure}{5.5cm}
\begin{tabular}{|lp{4.5cm}|}
\hline
\multicolumn{2}{|c|}{\Response}\\
\hline
\labelR:& Label identifying the request corresponding to this response\\
\timeR:& Natural number representing global order of (processing) requests\\
\dataR:& Data returned by the memory sub-system for a \Ld{} request\\
\hline
\end{tabular}
\end{subfigure}
\caption{Data types for request and response}
\label{req-resp}
\end{figure}

In the following, a \Response{} set only contains those responses sent by the
memory sub-system.

The responses should obey the following sanity conditions:

\begin{defn} (The labels in the responses are unique. Since \dataR{} field is
irrelevant for store requests, so uniqueness of responses does not depend on
\dataR{} field for stores.)
\small
\begin{multline*}
\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response),\\
r_1.\labelR = r_2.\labelR @-> r_1.\timeR = r_2.\timeR \wedge\\
(\reqFn(r_1.\labelR) = \Ld @-> r_1.\dataR = r_2.\dataR)
\end{multline*}
\end{defn}

\begin{defn} (No response can share the same time as another store response.)
\small
\begin{multline*}
\textit{uniqStTimes} := 
\forall (r_1, r_2 \in \Response),
r_1.\timeR = r_2.\timeR @-> \\
(\reqFn (r_1.\labelR)).\addrQ = (\reqFn (r_2.\labelR)).\addrQ @->\\
(\reqFn (r_1.\labelR)).\desc = \St @->
r_1.\labelR = r_2.\labelR
\end{multline*}
\end{defn}

@@

This brings us to the main condition for store atomicity which states that the
value returned for a load response corresponds to the \emph{last earlier} store
request that has been processed in the system for that address This is
indicated by the proposition \textit{lastEarlierValue}$(r)$ for response $r$.
The \emph{last} and \emph{earlier} are defined according to the field \timeR{}
returned in the responses. In case no store has happened earlier for a
particular address when a load is processed, then the load response returns the
initial value for the address. This is indicated by the proposition
\textit{initialValue}$(r)$ for response $r$.
\begin{defn}
\small
\begin{multline*}
\textit{initialValue} := \lambda (r \in \Response) \Rightarrow
(\reqFn (r.\labelR)).\desc = \Ld @->\\ r.\dataR = \initData (r.\addrQ) \wedge 
\forall (r' \in \Response), 0 \le r'.\timeR < r.\timeR @->\\
\neg ((\reqFn (r'.\labelR)).\addrQ = (\reqFn (r.\labelR)).\addrQ \wedge (\reqFn (r'.\labelR)).\desc = \St)
\end{multline*}
\end{defn}

\begin{defn}
\small
\begin{multline*}
\textit{initialValue} := \lambda (r \in \Response) \Rightarrow
(\reqFn (r.\labelR)).\desc = \Ld @->
\exists (r_m \in \Response),\\ r.\dataR = (\reqFn (r_m.\labelR)).\dataQ \wedge 
(\reqFn (r_m.\labelR)).\desc = \St \\ \wedge (\reqFn (r_m.\labelR)).\addrQ = (\reqFn (r.\labelR)).\addrQ\\ \wedge
\forall (r' \in \Response), r_m.\timeR \le r'.\timeR < r.\timeR @->\\
\neg ((\reqFn (r'.\labelR)).\addrQ = (\reqFn (r.\labelR)).\addrQ \wedge (\reqFn (r'.\labelR)).\desc = \St)
\end{multline*}
\end{defn}

Using the above definitions, we can define the store-atomicity theorem as follows:
\begin{thm}
\small
\begin{multline*}
storeAtomicity := 
\textit{uniqRespLabels} \wedge
\textit{uniqStTimes} \wedge \textit{localOrdering}\\ \wedge 
(\forall (r \in \Response), \textit{initialValue}(r) \vee \textit{lastEarlierValue}(r))
\end{multline*}
\end{thm}

>>>>

\section{Correctness of a Cache Coherence Protocol}
\label{Background}

Before we can discuss our protocol and proof framework we must first establish
exactly what it means for a cache coherence protocol to be correct.

Intuitively, the simplest view of shared memory is as a single ported memory
without caches multiplexed amongst the many individual clients/processors. In
this view, all operations in the system appear to have a single unique total
ordering.

All modern systems have caches between the processor and memory for faster
memory access.  It is often also useful to permit some reorderings of requests
from a processor to the memory to improve performance.  For instance, in the
context of a processor with local store buffer, we may wish to allow a store to
bypass a subsequent load request to a different location. The specification of
which reorderings are permitted or restricted defines the memory consistency
model.

Most memory consistency models can be partitioned into two
parts~\cite{Arvind-memory-model}: a) a ``global'' store-atomicity property
stating that for each location/address in memory, every read and write access
to that location occurs in some total order with each operation seeing the
effects of previous operations and not later ones (in other words, each
location behaves like an atomic register) and b), a ``local'' reordering
restriction between requests to different locations within each processor.
Changing the memory model by modifying the reordering restrictions will not
incur a reverification of the store-atomicity property.

The mechanism employed by the memory sub-system to ensure store-atomicity per
location, in the presence of caches, is known as \emph{cache coherence}. In the
presence of caches, multiple copies of the same address can exist in different
caches. If these copies are independently written, the same address can end up
having different values in different caches, violating the atomic register-like
behavior unless the caches co-ordinate via a \emph{cache coherence protocol}.

We will now provide the formal definition of store atomicity.

\subsection{Formal definition of store atomicity}
We will model an abstract memory sub-system, which interfaces with a set of
processors. The requests issued by the processors to the memory sub-system can
be modeled using a function \reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request
issued by processor $p$.

%Figure~\ref{req-resp} shows the
%data-type corresponding to a request issued from the processor to the
%memory sub-system.

A request contains the address (\addrQ) corresponding to the request, and a
\desc{} indicating whether the request is a load or a store. A store request
also contains the data supplied by the processor for that address (\dataQ). A
response contains a label (\labelR) containing a processor name and a position
to uniquely identify the request.  In the case of a load request, the memory
sub-system returns \dataR, the data returned by the memory sub-system for that
address. The memory sub-system also returns an $\mathbb{N}$, \timeR, which
intuitively represents the position at which the request is handled among the
global set of handled requests from all the processors. \timeR{} is used solely
for the definition of store atomicity, a real system does not return such a
field.

@@

%\begin{figure}
%\centering
%\begin{subfigure}{5.5cm}
%\begin{tabular}{|lp{4.5cm}|}
%\hline
%\multicolumn{2}{|c|}{\Request}\\
%\hline
%\addrQ:& Address associated with the request\\
%\desc:& Request type (\Ld{} or \St)\\
%\dataQ:& Data associated with a \St{} request\\
%\hline
%\end{tabular}
%\end{subfigure}~~~~
%\begin{subfigure}{5.5cm}
%\begin{tabular}{|lp{4.5cm}|}
%\hline
%\multicolumn{2}{|c|}{\Response}\\
%\hline
%\labelR:& Label identifying the request corresponding to this response\\
%\timeR:& Natural number representing global order of (processing) requests\\
%\dataR:& Data returned by the memory sub-system for a \Ld{} request\\
%\hline
%\end{tabular}
%\end{subfigure}
%\caption{Data types for request and response}
%\label{req-resp}
%\end{figure}

The \Response{} set contains the set of responses from the memory sub-system.
Responses should obey the following sanity conditions. 
\begin{defn} (The labels in the responses are unique)
%Since \dataR{} field is
%irrelevant for store requests, so uniqueness of responses does not depend on
%\dataR{} field for stores.)
\small
\begin{multline*}
\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response),\\
r_1.\labelR = r_2.\labelR @-> r_1.\timeR = r_2.\timeR \wedge\\
(\reqFn(r_1.\labelR) = \Ld @-> r_1.\dataR = r_2.\dataR)
\end{multline*}
\end{defn}

\begin{defn} (No response can share a \timeR{} value with another store response.)
\small
\begin{multline*}
\textit{uniqStTimes} := 
\forall (r_1, r_2 \in \Response),
r_1.\timeR = r_2.\timeR @-> \\
(\reqFn (r_1.\labelR)).\addrQ = (\reqFn (r_2.\labelR)).\addrQ @->\\
(\reqFn (r_1.\labelR)).\desc = \St @->
r_1.\labelR = r_2.\labelR
\end{multline*}
\end{defn}

@@

This brings us to the main condition for store atomicity which states that the
data returned for a load response corresponds to the \emph{last earlier} store
request from any processor that has been handled for that address,
\emph{last} and \emph{earlier} being defined according to the field \timeR{}
returned in the responses.  This proposition is indicated by
\textit{lastEarlierValue}$(a, t, d)$ for a load request for address $a$ handled
at time $t$, returning data $d$.  In case no store has happened earlier for a
particular address when a load is processed, then the load response returns the
initial value for the address, the whole proposition indicated by
\textit{initialValue}$(a, t, d)$.
\begin{defn}
\small
\begin{multline*}
\textit{initialValue} := \lambda (a \in Addr)\; (t \in \mathbb{N}) \; (d \in Data)\Rightarrow
d = \initData (a) \wedge \\
\forall (r' \in \Response), 0 \le r'.\timeR < t @->\\
\neg ((\reqFn (r'.\labelR)).\addrQ = t \wedge (\reqFn (r'.\labelR)).\desc = \St)
\end{multline*}
\end{defn}

\begin{defn}
\small
\begin{multline*}
\textit{lastEarlierValue} := \lambda (a \in Addr)\; (t \in \mathbb{N}) \; (d \in Data)\Rightarrow
\exists (r_m \in \Response),\\ d = (\reqFn (r_m.\labelR)).\dataQ \wedge 
(\reqFn (r_m.\labelR)).\desc = \St \wedge\\ (\reqFn (r_m.\labelR)).\addrQ = a \wedge
\forall (r' \in \Response), r_m.\timeR \le r'.\timeR < t @->\\
\neg ((\reqFn (r'.\labelR)).\addrQ = a \wedge (\reqFn (r'.\labelR)).\desc = \St)
\end{multline*}
\end{defn}

Using the above definitions, we can define the store-atomicity theorem as follows:
\begin{thm}
\small
\begin{multline*}
storeAtomicity := 
\textit{uniqRespLabels} \wedge
\textit{uniqStTimes} \wedge \textit{localOrdering} \wedge \\
(\forall (r \in \Response), \mylet q := \reqFn(r.\labelR) \myin q.\desc = \Ld \wedge \\
(\textit{initialValue}(q.\addrQ, r.\timeR, r.\dataR) \vee \textit{lastEarlierValue}(q.\addrQ, r.\timeR, r.\dataR)))
\end{multline*}
\end{thm}
